---
title: "CTMC Function Documentation and testing"
output: html_notebook
---

This script simulates the habitat dynamics that occur as a result of habitat patches being formed and lost (at a number of, "n", specified locations of "potential habitat" within a landscape) as a result of a disturbance regime dictated by a rate at which patches form ("r.on") and a rate at which they are lost ("r.off") from those locations. Thus, over time different habitat configurations arise based on which locations in the landscape host patches happen to be present (on) or absent (off) (represented by a binary vector "config" which indicates which patches are present (on) or absent (off) with 1 or 0 respectively), where transitions between these habitat configurations occur according to a continuous time markov chain (CTMC).

The simulation uses a Gillespie algorithm to optimize computation, since the time between changes "tau" (measuring time between changes in habitat for our CTMC) is exponentially distributed with mean 1/(sum of the rates of change for each patch). Thus, simulation is sped up by simulating only these changes at times they are expected to occur (draws from this exponential distribution) rather than what happens at every possible timestep over a time interval. This simulation is performed on an interval of zero to a maximum number of timesteps specified ("timesteps") or until it reaches the absorbing state (where no habitat remains, "config" = 0 for all patches).

NOTE: this function is actually very fast provided the expected time to absorption is low!

***It follows the following Algorithm:***

**1) Set up the patches in the initial configuration:**
Set up which patches are on or off according to the initial configuration given by the binary vector "initial.config" (1=on, 0=off)

**2) End if we began in the absorbing state:**
If all the patches were off in the initial configuration, then we are already in the absorbing state and the simulation ends

**3) Otherwise, (provided we are not in the absorbing state):**
Otherwise, for each timestep (provided at least one patch remains on and therefore we haven't entered the absorbing state and ended the simulation)...

  **3.1) Set the rates of transition for each patch**
  Set the rate each patch should turn on or off at according to whether they are on or off. Patches that are on turn off at rate "r.off" and patches that are off turn on at rate "r.on".

  **3.2) Calculate the time to the next transition**
  Calculate the time until the next configuration transition should occur (tau). For a CTMC tau is exponentially distributed with mean 1/(the sum of all the rates of transition)

  **3.3) Record the time between each transition**
  We would like to keep track of the time between each transition takes place at. We do so in a vector called tau.times. NOTE: We don't know how long this vector needs to be and so to be safe this vector is just being appended with new values each time (can slow a program down/eat memory -though memory not an issue on the server I'll run on). Probably it will be shorter than timesteps but because this is a contininuous time model though multiple transitions don't happen "at once" the time to the next transition could occur within the same integer timestep for which we are recording events.

  **3.4) Update the time elapsed in the simulation** 
  Update the time in the simulation to when this transition should occur (that is current t+tau)

  **3.5) Make the transition**
  Pick a random number between 1 and 0. Then break the range from 0 to 1 into intervals proportional to the probability each path should turn on or off.
  ie. for a config (0 1 0) the rate.vec is (r.on, r.off, r.on) so intervals proportional to the probability each patch should turn on or off can be defined by cumulative sum of the proportions of time each patch should turn on or off (r.on/(2r.on+r.off), (r.on/(2r.on+r.off))+(r.off/(2r.on+r.off)),(r.on/(2r.on+r.off))+(r.off/(2r.on+r.off))+r.on/(2r.on+r.off))
  Print the new configuration so we can check A) one patch that was either on or off turned either on or off and (because only one change can occur at a instaneous point in time -by definition of a CTMC) B) that the simulation is running through the transitions smoothly and timely
  
***List of Parameters***
* timesteps = max time we want to simulate over (note: slight misnomer: changes due not occur on discrete timestep intervals but rather at any time between 0 and timesteps)

* n.patches = number of locations at which habitat may be formed or 

* r.on = rate at which habitat is formed at a location

* r.off = rate at which habitat is lost at a location

* initial.config = a vector containing a 1 or 0 for each location where a patch is present (1) or absent (0) to start, where the first entry corresponds to location 1, 2nd to location 2 and so on

```{r}
CTMC.sim<-function(timesteps, n.patches, r.on, r.off, initial.config) {
  
  #1. SET UP THE PATCHES IN THE INITIAL CONFIGURATION
  #############################################################################################
  config<-initial.config #setting up patches in the initial configuration
  tau.times<-0 #no initial time until transition
  #############################################################################################
  
  #2. END IF WE BEGAN IN THE ABSORBING STATE
  #############################################################################################
  if (sum(config)==0){ #if we are already in the absorbing state
    end.t<-0 #time to absorption is 0
  #############################################################################################
    
    #3. OTHERWISE, PROVIDED WE'RE STILL NOT IN THE ABSORBING STATE...
    #############################################################################################
  } else { #otherwise
    for (t in 0:timesteps) { #comment out this line and end bracket if don't want to set a time limit
      #***actually can't do that in R
      while (sum(config)>0) { #while at least one patch remains on
        
        #3.1 SET THE RATES OF TRANSITION FOR EACH PATCH
        ###############################################
        rate.vec<-rep(NA, n.patches) #construct a vector of transition rates 
        #depending on initial configuration
        rate.vec[config==0]<-r.on #if the patch is off set it to turn on at rate r.on
        rate.vec[config==1]<-r.off #if the patch is on set it to turn off at rate r.off
        ###############################################
        
        #3.2 CALCULATE THE TIME TO THE NEXT TRANSITION
        #select the time until transition (exponential, with mean 1/rate.sum)
        #####################################################################
        rate.sum<-sum(rate.vec)
        tau<-rexp(1, rate.sum) #tau = time until next transition
        #####################################################################
        
        #3.3 RECORD THE TIME OF EACH TRANSITION
        #note: not necessary for simulation but useful to know
        ######################################################
        if (sum(tau.times) == 0){
          tau.times<-tau} else{
          tau.times<-c(tau.times, tau)}
        ######################################################
        
        #3.4 UPDATE THE TIME ELAPSED IN SIMULATION
        ##########################################
        t<-t+tau
        ##########################################
        
        #3.5 MAKE THE TRANSITION
        ########################
        cum.rate<-cumsum(rate.vec/rate.sum) #calculate the cumulative rate of transition
        #define intervals between 0 and 1 by the proportion of time each patch should transition
        upper<-cum.rate
        lower<-c(0,cum.rate[-n.patches])
        unif.val<-runif(1, min=0, max=1) #pick a random number between 0 and 1
        if(config[unif.val<=upper & unif.val>lower]==1){ #if the patch lying on the selected interval 
          #within which the random value fell, is on
          config[unif.val<=upper & unif.val>lower]<-0 #turn it off
        } else {config[unif.val<=upper & unif.val>lower]<-1} #otherwise, turn it on
        print(config)
        #######################
        
        end.t<-t #record the time to absorption or simulation end
      } #end of while loop
    } #end of for loop
  } #end of else bracket
 #############################################################################################
  return(list(end.t=end.t, tau.times=tau.times))
}
```


**TESTING**
Things to look for to ensure it is running correctly
A) only 1 transition occurs at a time
B) runs quickly and smoothly
C) sum of tau.times = end.t
```{r, eval=FALSE}
#INITIALIZING PARAMETERS
timesteps<-1000000 #total number of iterations to run the model (optional, if unspecified will run until absorption)
n.patches<-2 #number of patches in the landscape
r.on<-100 #rate at which patches turn from off to on
r.off<-1 #rate at which patches turn from on to off
initial.config<-rbinom(n.patches, 1, 0.5) #initial habitat configuration of patches on and off
#drawn from the binomial distribution with 0.5 probability of success getting 1 (or 0)
```

```{r, eval=False}
out<-CTMC.sim(1000, 2, r.on=100, r.off=1, initial.config = c(1,1))
out
```

```{r, eval=FALSE}
out$end.t
out$tau.times
sum(out$tau.times)
```

For further documentation and testing see the two patch simulation test case file "2 patch test case".

